Structured Synchronous Reactive Programming for Game Development \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}

### 2017-05-28 paper.tex
...

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control-flow mechanisms, such
      as \code{await} (to suspend a line of execution), and \code{par} (to
      combine multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution.
      There's no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU provides primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \emph{Finite State Machines}:
    Event occurrences lead to transitions between states and trigger actions
    comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation in the
    game, i.e., some action to execute next.
\item \emph{Dispatching Hierarchies}:
    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.
%\item \emph{Signaling Mechanisms}:
    %Entities often need to communicate and affect each other explicitly through
    %signaling mechanisms, especially if there is no hierarchy relationship
    %between them.
\end{enumerate}

@@

void Bomber::draw (SceneContext& gc) {
  // 3. particles, terrain, explosion sprite
  // 4. tick: hides the explosion sprite
  if (sprite.frame()==13 && !gfx_exploded) {
    gfx_exploded = true;
    gc.color().draw(explo_surf, <...>);
  }
  gc.color().draw(sprite, pingu->get_pos());
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName
do
  <...>
  spawn Mover(); // movement in the background
  var Sprite sprite = spawn Sprite(<...>);
                 // animation in the background
  watching s do
    // 1. plays a "Oh no!" sound.
    {play_sound("ohno")};

@@

    The completion of a long-lasting activity may carry a continuation in the
    game, i.e., some action to execute next.

@@

void StoryScreenComp::next_text() {
  if (!displayed) {
    <...>
  } else {
    <...>
    if (!pages.empty()) {
      <...>
    } else {
      if (credits) {
        replace_screen(<Credits>(<...>));
      } else {
        pop_screen();
      }
    }
  }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
loop do
  var int ret = await Worldmap();
  if ret==STORY_MAP or ret==STORY_CREDITS then
    <...>
    var bool is_click = await Story();
    if is_click and ret==STORY_CREDITS then
      <...>
      await Credits();
    end
  else
    <...>
  end
end

@@

The campaign world map has clickable blue dots for introductory and ending
ambience stories in the two extremes of the map progress trail.
For introductory stories, the game returns to the world map after displaying 
the story pages.
For ending stories, the game also displays a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) reads
the level file (ln. 5) to check whether its an ending story and should, after
termination, display the \emph{Credits screen}.
%
The boolean variable \code{credits} is passed to the class
\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
to do after displaying the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
%
When the method \code{next\_text} has no pages left to display (ln. 32--38),
it decides where to go next, depending on the continuation flag
\code{credits} (ln. 33--37).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to display as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
the player interact with it until a dot is clicked.
If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
and await its termination (ln. 5).
Finally, we check the returned values (ln. 6) to perhaps display the
\code{Credits} screen (ln. 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ uses an explicit stack to push the \code{Worldmap} screen.
    \item \CEU invokes the \code{WorldMap} screen expecting a return value
          (ln. 2).
    \end{itemize}
\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pushes the \code{Story} screen passing the continuation flag
          (ln. 10).
    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
          (ln. 2,5).
    \end{itemize}
\item \code{Story} $\longrightarrow$ \code{Credits}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ replaces the current \code{Story} screen with the \code{Credits}
          screen (ln. 34).
    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
          returns (ln. 8).
    \end{itemize}
\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
          screen.
          \CEU uses an enclosing \code{loop} to restart the process (ln. 1--13).
    \end{itemize}
\end{enumerate}

In contrast with C++, the screens in \CEU are decoupled and only the
\code{Main Loop} touches them: the \code{Worldmap} has no references to
\code{Story}, which has no references to \code{Credits}.

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay.

@@

TODO: code for Sprite update?

In C++, the class \code{Bomber} presented in Figure~\ref{lst.hier}.a declares a
\code{sprite} member (ln. 3) to handle its animation frames
(Figure~\ref{fig.bomber.fsm}).
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls \code{update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities, such as all pingus and traps, with the common interface
      \code{WorldObj} (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change between actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
The \code{draw} callback flows through the same dispatching hierarchy until
reaching the \code{Sprite} class.

@@

Note that dispatching hierarchies cross game engine code, suggesting that most
games also rely heavily on this control-flow pattern.
In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
them from 515 to 173 LoCs (not shown in Figure~\ref{tab.tree}), mostly due to
dispatching code removal.

@@

Figure~\ref{fig.ui} shows the game UI widgets with the action buttons, score
counters, and a small map which all coexist with the game screen during its
whole lifespan.

@@

In \CEU, the UI entities that coexist just have to be created in the same
lexical block in Figure~\ref{lst.ui}.b (ln. 3--5).
%
Since abstractions can react independently, they do not require a dispatching
container.
%
Lexical lifespan never requires containers, allocation and deallocation, or
explicit references.
In addition, all required memory is known at compile time, similarly to
stack-allocated local variables.
%
The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
lexical scope to delimit the lifespan of the explosion sprite to a single
frame (Figure~\ref{lst.bomber}, ln. 19--25).

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude%
\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }.
%

@@

void PinguHolder::update() {
    <...>
    while(pingu != pingus.end()) {
        (*pingu)->update();
        if ((*pingu)->get_status() == PS_DEAD) {
            pingu = pingus.erase(pingu);
        }
        <...>
        ++pingu;
    }
}

@@

In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
container that holds all alive pingus.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Entities with dynamic lifespan in C++ require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).
Without the \code{erase} call above, a dead pingu would remain in the
collection still with updates on every frame (ln. 11).
Since the \code{draw} behavior for a dead pingu is innocuous, the death could
go unnoticed when testing it but the program would keep consuming memory and
CPU time.
This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
also occurs in languages with garbage collection:
a container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
Additionally, the \code{spawn} statement supports a pool identifier to
associate the new instance with a pool.
%
The game screen in Figure~\ref{lst.pingus}.b spawns a new \code{Pingu} on every
invocation of \code{Pingu\_Spawn} (ln. 4--7).
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

@@

A number of domain-specific languages, frameworks, and techniques have been
proposed for particular subsystems of the game logic, such as
animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996,games.anims.1982},
game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
behavior and AI modeling~\cite{games.bts,games.bts.unreal}
%
In Pingus, we employed \CEU at the core of the game for event dispatching
(Section~\ref{sec.pats.dispatching}) and memory management of entities
(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
engine.
We also implemented all entity animations and behaviors
(Section~\ref{sec.pats.fsms}), and screen transitions
(Section~\ref{sec.pats.cps})
%, and intermodule communication (Section~\ref{sec.pats.signaling})
using the available control mechanisms of \CEU.
%
Furthermore, \CEU is a superset of C targeting reactive systems in general, not
only games, and has also been successfully adopted in other domains, such as
    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
    multimedia systems~\cite{ceu.media.webmedia16}.

Functional reactive programming (FRP)~\cite{frp.fran} contrasts with
structured synchronous reactive programming (SSRP) as a complementary
programming style for reactive applications.
%
We believe that FRP is more suitable for data-intensive applications, while 
SSRP, for control-intensive applications.
%
On the one hand, FRP uses declarative formulas to specify continuous functions 
over time, such as for physics or data constraints among entities.
%
On the other hand, describing a sequence of steps in FRP requires to encode 
explicit state machines so that functions can switch behavior depending on the 
current state.
%
FRP has been successfully used to implement a 3D first person shooting game
from scratch, but with performance considerations~\cite{games.frag}.
%
Instead, we rewrote an existing game and did it in small steps while keeping it
working.
Although we do not provide a performance evaluation (Pingus is not performance
sensitive), previous work on \CEU shows that it is comparable to C in the
context of embedded systems~\cite{ceu.sensys13}.
Nonetheless, given the tight integration between with C/C++, critical parts of
games can be preserved in C++ if needed.

>>>>

In this work, we advocate structured synchronous reactive programming as an
expressive and productive alternative for game logic development.
We present a case study of rewriting Pingus from C++ to the programming
language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
characteristics:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control-flow mechanisms, such
      as \code{await} (to suspend a line of execution), and \code{par} (to
      combine multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution.
      There is no implicit preemption or real parallelism, resulting in
      deterministic execution.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU provides primitives that help describing complex control-flow
relationships in the game logic more concisely.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and does not rely on garbage collection.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \emph{Finite State Machines}:
    Event occurrences lead to transitions between states and trigger actions
    comprising the behavior of a game entity.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity in the game may carry a
    continuation, i.e., some action to execute next.
\item \emph{Dispatching Hierarchies}:
    Entities form a dispatching hierarchy in which a container that receives a
    stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities form a lifespan hierarchy in which a terminating container entity
    automatically destroys its managed children.
%\item \emph{Signaling Mechanisms}:
    %Entities often need to communicate and affect each other explicitly through
    %signaling mechanisms, especially if there is no hierarchy relationship
    %between them.
\end{enumerate}

@@

void Bomber::draw (SceneContext& gc) {
  // 3. particles, terrain, explosion sprite
  // 4. tick: hides the explosion sprite
  if (sprite.frame()==13 && !gfx_exploded) {
    gfx_exploded = true;
    gc.color().draw(explo_surf, <...>);
  }
  gc.color().draw(sprite, pingu->get_pos());
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Bomber (void) -> ActionName
do
  <...>
  spawn Mover(); // movement in background
  var Sprite sprite = spawn Sprite(<...>);
               // frame animation in background
  watching sprite do
    // 1. plays a "Oh no!" sound.
    {play_sound("ohno")};

@@

    The completion of a long-lasting activity in the game may carry a
    continuation, i.e., some action to execute next.

@@

void StoryScreenComp::next_text() {
  if (!displayed) {
    <...>
  } else {
    <...>
    if (!pages.empty()) {
      <...>
    } else {
      if (credits) {
        replace_screen(<Credits>(<...>));
      } else {
        pop_screen();
      }
    }
  }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
loop do
  var int ret = await Worldmap();
  if ret==STORY_CREDITS or ret==STORY_BACK then
    <...>
    var bool is_click = await Story();
    if is_click and ret==STORY_CREDITS then
      <...>
      await Credits();
    end
  else
    <...>
  end
end

@@

The campaign world map has clickable blue dots in the two extremes of the
progress trail for introductory and ending ambience stories, respectively.
For introductory stories, the game returns to the world map after displaying 
the story pages.
For ending stories, the game also displays a \emph{Credits screen} before
returning to the world map.%
\footnote{Credits screen animation: \url{github.com/an000/p/#4} }

In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
reads the level file (ln. 5) to check whether its an ending story and should,
after termination, display the \emph{Credits screen}.
%
The boolean variable \code{credits} is passed to the class
\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
to do after displaying the story.
The class \code{StoryScreen} (not shown) then forwards the continuation even
further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
%
When the method \code{next\_text} has no pages left to display (ln. 32--38),
it decides where to go next, depending on the continuation flag
\code{credits} (ln. 33).

In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
the screens to display as a direct sequence of statements.
%
We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
the player interact with it (e.g., walking around) until a dot is clicked.
If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
and await its termination (ln. 5).
Finally, we check the returned values (ln. 6) to perhaps display the
\code{Credits} screen (ln. 8).
The enclosing loop restores the \code{Worldmap} and repeats the process.

@@

\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ uses an explicit stack to push the \code{Worldmap} screen.
    \item \CEU invokes the \code{Worldmap} screen expecting a return value
          (ln. 2).
    \end{itemize}
\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pushes the \code{Story} screen passing the continuation flag
          (ln. 10).
    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
          (ln. 2,5).
    \end{itemize}
\item \code{Story} $\longrightarrow$ \code{Credits}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ replaces the current \code{Story} screen with the \code{Credits}
          screen (ln. 34).
    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
          returns (ln. 8).
    \end{itemize}
\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
    \begin{itemize}
    %\setlength\itemsep{0em}
    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
          screen.
          \CEU uses an enclosing \code{loop} to restart the process (ln. 1--13).
    \end{itemize}
\end{enumerate}

In contrast with C++, the screens in \CEU are completely decoupled and only the
\code{Main Loop} touches them: the \code{Worldmap} has no references to
\code{Story}, which has no references to \code{Credits}.

@@

Continuation passing typically controls the overall structure of the game,
such as screen transitions in menus and level progressions.
%
\CEU uses the direct style techniques in five cases involving screen
transitions:
the main menu, the level menu, the level set menu, the world map loop, and
the gameplay loop.
%
It also uses the same technique for the loop that switches the pingu actions
during gameplay (e.g., \emph{walking} to \emph{jumping} and back to
\emph{walking}).

@@

In C++, the class \code{Bomber} in Figure~\ref{lst.hier}.a declares a
\code{sprite} member (ln. 3) to handle its animation frames.
%
The \code{Sprite} class is part of the game engine and knows how to update and
render itself.
However, the \code{Bomber} still has to respond to \code{update} and
\code{draw} requests from the game and forward them to the sprite
(ln. 11--13 and 15--18).
%
To understand how the \code{update} callback flows from the original
environment stimulus to the game down to the sprite, we need to follow a long
chain of 7 method dispatches (Figure~\ref{fig.hier}):
%
\begin{enumerate}
%\setlength\itemsep{0em}
\item \code{ScreenManager::display} in the main game loop calls
      \code{ScreenManager::update}.
\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
      active game screen (i.e., a \code{GameSession} instance, considering the
      \code{Bomber}).
\item \code{GameSession::update} calls \code{world->update}.
\item \code{World::update} calls \code{obj->update} for each object in the
      world.
\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
      alive.
\item \code{Pingu::update} calls \code{action->update} for the active pingu
      action.
\item \code{Bomber::update} calls \code{sprite.update}.
      \code{Sprite::update} finally updates the animation frame.
\end{enumerate}
%
Each dispatching step in the chain is necessary considering the game
architecture:
%
\begin{itemize}
%\setlength\itemsep{0em}
\item With a single assignment to \code{last\_screen}, we can easily deactivate
      the current screen and redirect all dispatches to a new screen (step 2).
\item The \code{World} class manages and dispatches events to all game
      entities with a common interface \code{WorldObj}, such as all pingus and
      traps (step 4).
\item Since it is common to iterate only over the pingus (vs. all world
      objects), the container \code{PinguHolder} manages all pingus (step 5).
\item Since a single pingu can change its actions during lifetime, the
      \code{action} member decouples them with another level of indirection
      (step 6).
\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
      buttons, world objects, etc.), so it is also convenient to decouple them
      from actions (step 7).
\end{itemize}
%
The \code{draw} callback also flows through a similar dispatching hierarchy
until reaching the \code{Sprite} class.

@@

Note that dispatching hierarchies cross game engine code, suggesting that most
games also rely heavily on this control-flow pattern.
In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
them from 515 to 173 LoCs (not listed in Figure~\ref{tab.tree}), mostly due to
dispatching code removal.

@@

Figure~\ref{fig.ui} shows the game UI widgets with action buttons, score
counters, and a small map which all coexist with the game screen during its
whole lifespan.

@@

In \CEU, the UI entities that coexist are just created in the same lexical
block of the abstraction \code{Game} in Figure~\ref{lst.ui}.b (ln. 3--5).
%
Since abstractions can react independently, they do not require a dispatching
container.
%
Lexical lifespan never requires containers, allocation and deallocation, or
explicit references.
In addition, all required memory is known at compile time, similarly to
stack-allocated local variables.
%
The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
lexical scope to delimit the lifespan of the explosion sprite to a single
frame (Figure~\ref{lst.bomber}, ln. 19--25).

@@

A pingu is a dynamic entity created periodically and destroyed under certain
conditions, such as falling from a high altitude.%
\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }
%

@@

void PinguHolder::update() {
    <...>
    while(pingu != pingus.end()) {
        (*pingu)->update();
        if ((*pingu)->get_status() == PS_DEAD) {
            pingu = pingus.remove(pingu);
        }
        <...>
        ++pingu;
    }
}

@@

In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
container that holds all alive pingus.
%
The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
to create a new \code{Pingu} and add it to the \code{pingus} collection
(ln. 3--4).
The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
pingus on every frame, removing those with the dead status (ln. 12--14).
%
Without the \code{remove} call, a dead pingu would remain in the collection
still with updates on every frame (ln. 11).
Since the \code{draw} behavior for a dead pingu is innocuous, the death could
go unnoticed when testing it but the program would keep consuming memory and
CPU time.
%
This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
also occurs in languages with garbage collection:
a container typically holds a strong reference to a child (sometimes the only 
reference to it), and the runtime cannot magically detect it as garbage.
%
Entities with dynamic lifespan in C++ always require explicit \code{add} and
\code{remove} calls associated to a container (ln. 4,13).

@@

\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
In addition, the \code{spawn} statement supports a pool identifier to associate
the new instance with a pool.
%
The game screen in Figure~\ref{lst.pingus}.b spawns a new \code{Pingu} on every
invocation of \code{Pingu\_Spawn} (ln. 4--7).
%
The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
block of the game screen (ln. 3).
In this case, the lifespan of the new instances follows the scope of the pool
(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
(ln. 4--7).
Since pools are also subject to lexical scope, the lifespan of all dynamically
allocated pingus is constrained to the game screen.
%
Lexical scopes handle memory and event dispatching automatically for static
instances and also for pools.
However, the lifespan of a dynamic instance does not necessarily have to match
the lifespan of its associated pool (Figure~\ref{fig.pool}).
In \CEU, when the execution block of a dynamic instance terminates, which
characterizes its \emph{natural termination}, the instance is automatically
removed from its pool.
Therefore, dynamic instances do not require any extra bookkeeping related to 
containers or explicit deallocation.
%
To remove a pingu from the game in \CEU, we just need to terminate its execution
block according to the appropriate conditions:
%
The \code{escape} statement (ln. 17) aborts the execution block of the
\code{Pingu} instance, removing it from its associated pool automatically.
Hence, a dynamic instance that terminates naturally leaves no traces in the 
program.

@@

A number of domain-specific languages, frameworks, and techniques have been
proposed for particular subsystems of the game logic, such as
animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996,games.anims.1982},
game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
behavior and AI modeling~\cite{games.bts,games.bts.unreal}.
%
In Pingus, the adoption of \CEU is not restricted to a specific subsystem.
We employed \CEU at the very core of the game for event dispatching
(Section~\ref{sec.pats.dispatching}) and memory management of entities
(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
engine.
We also implemented all entity animations and behaviors
(Section~\ref{sec.pats.fsms}), and screen transitions
(Section~\ref{sec.pats.cps})
%, and intermodule communication (Section~\ref{sec.pats.signaling})
using the available control mechanisms of \CEU.
%
Furthermore, \CEU is a superset of C targeting reactive systems in general, not
only games, and has also been successfully adopted in other domains, such as
    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
    multimedia systems~\cite{ceu.media.webmedia16}.

Functional reactive programming (FRP)~\cite{frp.fran} contrasts with
structured synchronous reactive programming (SSRP) as a complementary
programming style for reactive applications.
%
We believe that FRP is more suitable for data-intensive applications, while 
SSRP, for control-intensive applications.
%
On the one hand, FRP uses declarative formulas to specify continuous functions 
over time, such as for physics or data constraints among entities.
%
On the other hand, describing a sequence of steps or control-flow dependencies
in FRP requires to encode explicit state machines so that functions can switch
behavior depending on the current state.
%
FRP has been successfully used to implement a 3D first person shooting game
from scratch, but with performance considerations~\cite{games.frag}.
%
Instead, we rewrote an existing game and did it in small steps while keeping it
working.
Although we do not provide a performance evaluation (Pingus is not performance
sensitive), previous work on \CEU shows that it is comparable to C in the
context of embedded systems~\cite{ceu.sensys13}.
Nonetheless, given the tight integration between \CEU and C/C++, critical parts
of games can be preserved in C++ if needed.
